---
title: "Growth Rate"
author: "Kenji Sato"
date: 'Day 2'
output:
  xaringan::moon_reader:
    mathjax: "https://cdnjs.cloudflare.com/ajax/libs/mathjax/2.7.1/MathJax.js?config=TeX-AMS-MML_HTMLorMML"
    css: 
      - "../asset/remark-css/style.css"
    lib_dir: libs
    nature:
      countIncrementalSlides: no
      highlightLines: yes
      highlightStyle: github
subtitle: Macroeconomics (2018Q3)
header-includes: <script src="https://use.fontawesome.com/b5f301ecca.js"></script>
---


```{r setup, include=FALSE}
knitr::opts_chunk$set(fig.align = 'center')
source("../R/utils.R")
library(tidyverse)
library(cowplot)
```


## Growth Rate

```{r, include=FALSE}
Y = c(15591, 15978, 16274, 16705)
```

| Year  | GDP    | Symbol  |
|:-----:|:------:|:-------:|
| 2011  | 15,591 | $Y_1$   |
| 2012  | 15,978 | $Y_2$   |
| 2013  | 16,274 | $Y_3$   |
| 2014  | 16,705 | $Y_4$   |

Growth rate between years 2011 and 2012 is 

$$g_{2,1} = \frac{Y_2}{Y_1} - 1 \simeq `r Y[2] / Y[1] - 1`$$

---

## Average growth rate

The average annual growth rate can be computed by geometric average

$$\begin{aligned}
  \left(
    \frac{Y_4}{Y_1}
  \right)^{\frac{1}{4 - 1}} - 1
  \simeq
  `r (Y[4] / Y[1])^(1/(4 - 1)) - 1`
\end{aligned}$$

Notice that 

$$\begin{aligned}
  \left(
    \frac{Y_4}{Y_1}
  \right)^{\frac{1}{4 - 1}} - 1
  =
  \underbrace{\sqrt[3]{\frac{Y_4}{Y_3} \frac{Y_3}{Y_2} \frac{Y_2}{Y_1}}}_{\text{geometric mean}} - 1
\end{aligned}$$

You don't use arithmetic mean to compute average growth rate. 

---

## mean log difference

$$\begin{aligned}
  \frac{\ln Y_4 - \ln Y_1}{4 - 1}
  \simeq
  `r (log(Y[4]) - log(Y[1])) / (4 - 1)`
\end{aligned}$$

Difference in log, devided by the length of the period, 
is close to the average growth rate.

Geometric average corresponds to **effective rate**, while 
the mean log difference corresponds to **nominal rate**.

---

## Compounding

Assume that a bank offers an annual, nominal interest rate of 6% compounded monthly and that you make a deposit of one thousand dollars ($1,000) at the bank today. 

How much do you expect to have in the bank account in one year from now?

--

$$1000 \times 
\left(
  1 + \frac{0.06}{12}
\right)^{12}
\simeq
`r 1000 * (1 + 0.06 / 12) ^ 12`$$

The rate of return is about 6.17%, which is larger than 6%.

---

## Two annual interest rates

6% = Nominal annual interest rate  
6.17% = Effective annual interest rate

Be aware of the difference.

As far as differential equations are concerned, nominal rates are more convenient. 

---

`r section("Differential Equations")`

---

## More frequent compounding

Compounded $N$ times in one year.

$$\lim_{N\to \infty} \left(
  1 + \frac{0.06}{N}
\right)^{N} 
=
e^{0.06} = \exp(0.06)$$

Alternatively, let $\Delta = 1/N$ to get 

$$\lim_{\Delta t \to 0} \left(
  1 + 0.06 \Delta t
\right)^{1/\Delta t} 
=
e^{0.06} = \exp(0.06)$$

That is,

* power 0.06 = nominal rate  
* $e^{0.06}$ = effective rate

Of course, log effective rate is nominal rate: $\ln e^{0.06} = 0.06$


---

## Relationship between nominal and effective rates

$$\begin{aligned}
  \left(
    \frac{Y_T}{Y_0}
  \right)^\frac{1}{T}
  &=
  \exp \left(
    \ln \left(
    \frac{Y_T}{Y_0}
  \right)^\frac{1}{T}
  \right) \\
  &=
  \exp \left(
    \frac{\ln Y_T - \ln Y_0}{T}
  \right)
\end{aligned}$$

$\frac{\ln Y_T - \ln Y_0}{T}$ is the nominal rate.

---

## Constant nominal growth rate

The most important case is when the nominal growth rate is constant.

$$\begin{aligned}
  \frac{\ln Y_t - \ln Y_0}{t}
  =
  g
\end{aligned}$$

Rearranging, 

$$\ln Y_t = \ln Y_0 + gt$$

which tells you why semilog plots exhibit linearity. By taking log of both sides:

$$Y_t = Y_0 e^{gt}$$

---

## Continuous-time modeling

Let $t$ be any nonnegtive real number. 

To clarify that $Y$ is a function of time, we rewrite as $Y(t) = Y(0) e^{gt}$

```{r, echo=FALSE, out.width='350px'}
t <- seq(1, 70, 0.1)
Y <- function(x) exp(0.02 * x)
ggplot2::ggplot(tibble(t, Y = Y(t))) + geom_line(aes(t, Y)) +
  theme(text = element_text(size=30))
```

---

## Differential equation

Differentiate

$$Y(t) = Y(0) e^{gt}$$

with respect to time to get

$$\frac{dY}{dt} (t) = g Y(0) e^{gt} = gY(t).$$

For notational simplicity, we write the time-derivative as

$$\dot Y = \frac{dY}{dt}$$


---

## Differential equation (cont'd)

The following equation is an example of differential equations, in which
a function is the unknown.

$$\dot Y = gY \quad \text{or}\quad 
\frac{\dot Y}{Y} = g$$

This is the most important differential equation we use. 

How can we get $Y(t) = Y(0)e^{gt}$ from this differential equation? 

In other words, how can we **solve** this differential equation?


---

## Solving simple differential equation

Notice that

$$\frac{d}{dt}\left[ 
  \ln Y(t)
\right]
=
\frac{\dot Y(t)}{Y(t)}$$

We thus have

$$\frac{d}{dt}\left[ 
  \ln Y(t)
\right] = g
\Longrightarrow
\int d\left[
  \ln Y(t)
\right] = \int gdt$$

Then

$$\ln Y(t) = gt + \text{const.}$$

---

## Solving simple differential equation (cont'd)

At $t = 0$,

$$\ln Y(0) = \text{const.}$$

We therefore have

$$\ln Y(t) = \ln Y(0) + gt
\Longrightarrow
Y(t) = Y(0) e^{gt}$$

To sum up,

**
$$\frac{\dot Y}{Y} = g
\Longleftrightarrow
Y(t) = Y(0) e^{gt}$$
**

---

## Continuous-time modeling

We will use **differential equations** to model the evolution of 
the economy over time (i.e., dynamics of the model).

Why is this strategy helpful?

---

## Force driving the economy

$K(t)$ is the amount of physical capital at time $t$

* machines
* buildings
* etc.

Capital is used for current and future production and is a fundamental 
source of wealth of the economy. 

Consider an increase in $K(t)$ as one form of economic growth.

---

## Force driving the economy (cont'd)

**Why does $K(t)$ increase?**

**Why does $K(t)$ decrease?**



---

## Force driving the economy (cont'd)

**Why does $K(t)$ increase?** → Investment

* Let $I(t)$ denote the speed of investment ($/year) at time $t$.

**Why does $K(t)$ decrease?** → Depreciation

* Let $\delta$ denote the annual rate of depreciation. 
* $\delta K(t)$ is the speed of depreciation at time $t$.

The balance between the two forces determines the path of capital
accumulation.

---

## Investment

Let $K(t)$ be given. Let's observe how $K(t + \Delta t)$ is determined.

The speed of investment is $I(t)$. 

If the economy keeps that speed from $t$ to $t + \Delta t$, then
the increment of capital between the two dates would be 

$$I(t)\Delta t$$

---

## Capital depreciation

The annual rate of depreciation is $\delta$. 

In $\Delta t$ years, capital is reduced by 

$$(\delta \Delta t) K(t)$$

---

## Capital accumulation

Combining these forces,

$$K(t + \Delta t) - K(t) = I(t)\Delta t - (\delta \Delta t) K(t)$$

--

Deviding by $\Delta t$, we get

$$\frac{K(t + \Delta t) - K(t)}{\Delta t}
=
I(t) - \delta K(t)$$

--

By taking limit $\Delta t \to 0$,

$$\dot K(t) = I(t) - \delta K(t)$$

---

## Differential equation, too mathematical?

Instead of directly modeling the variable of interest $K$,  
we start by describing increment/decrement of $K$.

This modeling strategy gives a clear picture of how the economy
evolves over time.

* When $I(t) > \delta K(t)$, i.e., when investment exceeds depreciation, then 
  $\dot K > 0$ and thus  $K$ is growing.
* When $I(t) < \delta K(t)$, i.e., when depreciation exceeds investment, then
  $\dot K < 0$ and thus $K$ is decreasing.
* When $I(t) = \delta K(t)$, $K(t)$ stays steadily.




